DRAFT: The algebraic hierarchy in type theory using type classes

### 2010-02-26 jar.tex
Small changes

We abstract from concrete algebra and formalize many-sorted universal algebra, including a
construction of
free $\tau$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
In fact, our aim is to use the present development as a new basis for Bishop's program in type
theory~\cite{typesreal-article}.\marginpar{More}

@@

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
This type theory lacks quotient types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
As we will see in Section~\ref{univ2} below this functionality can also be used to work with
quotient algebras.

Future versions of Coq are planned to support proof irrelevance: every |p,q:P:Prop| are
convertible; see section~\cite{explit}.

\subsection{Type classes}
Implementation in Ltac, Notation etc\\
Coercions\\
Unification hints

@@

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: carrier → carrier → Prop): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: carrier → carrier → Prop): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter.}
we will return to this issue in Section~\ref{explicit}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: er_carrier → er_carrier → Prop
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

@@

This list will grow in subsequent sections.

@@

\section{Universal algebra}\label{univ}
After having implemented concrete algebraic structures, we generalize some of our constructions
by implementing many-sorted universal algebra. A formalization of universal algebra
already exists in Coq~\cite{DBLP:conf/tphol/Capretta99} and has recently been extended with
more categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of
this using type classes. We define equational theories and their varieties of algebras, i.e.\ their
finitary algebraic category of models. A quote function helps to translate a statement for an
instance of
a type class to a statement in the equational logic. We will come back to this in
Section~\ref{quote}.

@@

We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
theorem.~\marginpar{Description of the work.}
Discuss: the duplication of code which may have been avoided by universe
polymorphism for |Fixpoint|s.
%perhaps the soundness theorem for equational logic?

In order to prove the first homophism theorem, we define a congruence.
The compatibility of the relation |e| can be stated in two provably  equivalent ways.
Directly, |eAlgebra := @Algebra et v e _| or  indirectly, by stating that the relation
as a set of pairs is a subalgebra in the product algebra.
%theory/ua_congruence.v
% \begin{code}
%  eSub := @ua_subalgebraT.ClosedSubset et vv _ _ Q.
% \end{code}
Subsequently, we choose for former as a definition:
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
A congruence relation is coarser than the setoid relation, as indicated by the use of |Proper|
above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
congruences and quotients benefits from this.

@@

We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra and an
object in the category of $\\tau$-algebras. Conversely, given an object in the category/variety, we
construct a monoid.
%varieties/monoid.v
\begin{code}
Variable o: variety.Object theory.
Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
Global Instance: MonoidUnit (o tt) := algebra_op theory one.
Global Instance from_object: Monoid (o tt).
\end{code}
In the last line, instance resolution `automatically' finds the operation and the unit specified in
the lines before.\marginpar{Why this way?}

\subsection{Category theory}\label{cats}
Using type classes, we provide a basic library for category theory: categories, functors,
adjunctions, monads,\ldots We follow the usual type theoretical definition of a
category~\cite{saibi1995constructive}

%format ◎ = "\circ"
%interfaces/abstract_algebra.v
\begin{code}
Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
  `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z,
      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
  ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{code}
Where |%signature|
tells Coq how to interpret |equiv|.\marginpar{Where is this defined?}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

@@

We have defined the category of setoids, the category of categories, the dual of a category,
products, and hence co-products, the category of $\tau$-algebras.

@@

\section{Type class quoting}\label{quote}
Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
function\footnote{Gonthier provides similar functionality using canonical structures}.
Unification hints are absent in Coq. Fortunately, type classes provide similar functionality as
we will now show.

We define a term language for monoids:
\begin{code}
Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
\end{code}
This defines a monad on the category of |Type|s with extensional functions.\marginpar{Proof this
formally}.
The expression type is parameterized over the set of variable indices. Hence, we diverge
 from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
 dummy variables.

@@

\section{Subset types and proof irrelevance}
We have mentioned proof irrelevance a number of times before. It is an important theme in our
approach. The |Program| machinery allows one to conveniently write programs with Hoare-style
pre-
and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side conditions
are intended to be proof irrelevant, they are in |Prop|. Presently, Coq does not support such subset
types, thus forcing the user to manually prove that |f| does not depend on the proof of |P|.
The addition of proof irrelevance and subset types to Coq is pursued by
Barras~\cite{Barras:subset,Werner}.

% Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
% \ldots.

We add another principle:
\begin{enumerate}\setcounter{enumi}{3}
 \item Working hypothesis: every record can be reordered in such a way that all the non-informative
fields come last.
\end{enumerate}

Pollack~\cite{pollack2000dependently} introduces record types with telescopes. 
Given a telescope\[
T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_{k_1})]
\]
There is an inductively defined type $\Sigma T$, with a single constructor, given by the following 
formation and introduction rule in pseudo-code for $k=2$:
\begin{align*}
 A_1 :& Type\\
 A_2 :& \Pi A_1,Type\\
\cline{1-2}
\Sigma T :& Type
\end{align*}

@@

\subsection{Propositions as [ ]-Types}\label{explicit}
As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
them. In practice, it is convenient to assume that all proofs of a proposition are equal: the
principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
This allows us to extract the program we want, but we want more:
We would like to use Coq as a dependently typed programming language. To do so,
one would like to add the principle of proof irrelevance to Coq. A promissing proposal is the
Calculus of implicit constructions with implicit
$\Sigma$-types~\cite{miquel2001implicit,barras2008implicit,barrasSigma}. Such a change would fit
naturally with the Hoare-style separation between programs and proofs which is used by the |Program|
machinery.

One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. However,
there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
of Propositions as types seems necessary.

>>>>

We abstract from concrete algebra and formalize many-sorted universal algebra, including a
construction of free $\tau$-algebras; see Section~\ref{univ}.

% We are also concerned with efficiency.
% In fact, our aim is to use the present development as a new basis for Bishop's program in type
% theory~\cite{typesreal-article}.\marginpar{More}

@@

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
This type theory lacks quotient types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
As we will see in Section~\ref{univ} below this functionality can also be used to work with
quotient algebras.

Coq has two kinds of sorts |Prop| for propostions and |Type| for data types. Coq
is scheduled to support proof irrelevance: every |p,q:P:Prop| are
convertible in the future; see section~\ref{PI}.

% \subsection{Type classes}
% Implementation in Ltac, Notation etc\\
% Coercions\\
% Unification hints

@@

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: carrier → carrier → Prop): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: carrier → carrier → Prop): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter.}
we will return to this issue in Section~\ref{PI}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: er_carrier → er_carrier → Prop
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

@@

@@

\section{Universal algebra}\label{univ}
After having implemented concrete algebraic structures, we generalize some of our constructions
by implementing many-sorted universal algebra. A formalization of universal algebra
already exists in Coq~\cite{DBLP:conf/tphol/Capretta99} and has recently been extended with
more categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of
this using type classes. We define equational theories and their category of algebras. A quote
function helps to translate a statement for an instance of a type class to a statement in the
equational logic. We will come back to this in Section~\ref{quote}.

@@

We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
theorem.~\marginpar{Description of the work.}
Discuss: the duplication of code which may have been avoided by universe
polymorphism for |Fixpoint|s.\marginpar{universe polymorphism}
%perhaps the soundness theorem for equational logic?

In order to prove the first homophism theorem, we define a congruence.
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
Alternatively, we could have state the compatibility of the relation |e| by stating that the
relation as a set of pairs is a subalgebra in the product algebra. We prove that both approaches
are equivalent.

A congruence relation is coarser than the setoid relation, as indicated by the use of |Proper|
above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
congruences and quotients benefits from this.\marginpar{how}

@@

We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
of monoids. Given a monoid, we can construct the corresponding $\tau_m$-algebra and an
object in the category of $\tau$-algebras. Conversely, given an object in the category/variety, we
construct an instance of the |Monoid| type class.\marginpar{Update code}
%varieties/monoid.v
\begin{code}
Variable o: variety.Object theory.
Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
Global Instance: MonoidUnit (o tt) := algebra_op theory one.
Global Instance from_object: Monoid (o tt).
\end{code}
In the last line, instance resolution `automatically' finds the operation and the unit specified in
the lines before.\marginpar{Why this way?}

\subsection{Category theory}\label{cats}
Using type classes, we provide a basic library for category theory: categories, functors,
adjunctions, monads,\ldots We follow the usual type theoretical definition of a
category~\cite{saibi1995constructive}.\marginpar{Update code}

%format ◎ = "\circ"
%interfaces/abstract_algebra.v
\begin{code}
Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
  `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z,
      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
  ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{code}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

@@

We have defined the category of setoids, the category of categories, the dual of a category,
products, and hence co-products and the category of $\tau$-algebras, for a given type $\tau$.

@@

\section{Type class quoting}\label{quote}
Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
function\footnote{Gonthier provides similar functionality using canonical structures}. However,
this feature is absent from Coq. Fortunately, type classes provide similar functionality as
we will now show.

We define a term language for monoids:
\begin{code}
Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
\end{code}
This defines a monad on the category of |Type|s with extensional functions.\marginpar{Proof this
formally?}
The expression type is parameterized over the set of variable indices. Hence, we diverge
 from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
 dummy variables.

@@

\section{Subset types and proof irrelevance}\label{PI}
We have mentioned proof irrelevance a number of times before. It is an important theme in our
approach. The |Program| machinery allows one to conveniently write programs with Hoare-style
pre- and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side
conditions
are intended to be proof irrelevant, they are in |Prop|. Presently, Coq does not support such subset
types, thus forcing the user to manually prove that |f| does not depend on the proof of |P|.
The addition of proof irrelevance and subset types to Coq is pursued by
Barras~\cite{Barras:subset,Werner}. 
% Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
% \ldots.

Pollack~\cite{pollack2000dependently} introduces record types with telescopes. 
Given a telescope\[
T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_k)].
\]
There is an inductively defined type $\Sigma T$, with a single constructor, given by the following 
formation and introduction rule in pseudo-code for $k=2$:
\begin{align*}
 A_1 :& Type\\
 A_2 :& \Pi A_1,Type\\
\cline{1-2}
\Sigma T :& Type
\end{align*}

@@

\subsection{Propositions as [ ]-Types}
As originally concieved~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
them. In practice, it is convenient to assume that all proofs of a proposition are equal: the
principle of proof irrelevance. These two views clearly contradict eachother. In Coq, a middle way
was choosen, we are forbidden to extract information from proofs, but proofs are not identified.
This allows us to extract the program we want, but we want more:
We would like to use Coq as a dependently typed programming language. To do so,
one would like to add the principle of proof irrelevance to Coq. A promissing proposal is the
Calculus of implicit constructions with implicit
$\Sigma$-types~\cite{miquel2001implicit,barras2008implicit,barrasSigma}. Such a change would fit
naturally with the Hoare-style separation between programs and proofs which is used by the |Program|
machinery.

One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
the [ ]-types are isomorphic to the type of sort |Prop|. However,
there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
of Propositions as types seems necessary.
